---
title: "Pertussis and the CMI-PB project"
author: "Kristiana Wong A16281367"
format: pdf
---

##Investigating pertussis cases by year

FIrst we will examine and explore Pertussis case number in the US as tracked by the CDC: https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

Lets use the datapasta package to scrape the CDC data from the website into R
```{r}
library(datapasta)

cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
         No..Reported.Pertussis.Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
cdc
```

Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

Now lets plot this into ggplot to make a plot of case numbers over time
```{r}
library(ggplot2)

ggplot(cdc) +
  aes(cdc$Year, cdc$No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(x="Year", y="Number of Cases", title="Pertussis Cases by Year (1922-2021)") +
  geom_vline(xintercept= 1931, color = "blue", label = "First US Vaccine Developed") +
  geom_vline(xintercept= 1970, color = "red", label = "Pertussis Case Decline") +
  geom_vline(xintercept=2020, color = "gold", label = "COVID-19 Pandemic")
```
Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?
```{r}
library(ggplot2)

ggplot(cdc) +
  aes(cdc$Year, cdc$No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(x="Year", y="Number of Cases", title="Pertussis Cases by Year (1922-2021)") +
  geom_vline(xintercept= 1946, color = "blue") +
  geom_vline(xintercept= 1996, color = "red") 
```
After the aP vaccine introduction, there is actually a gradual increase in the number of cases versus after the introduction of the wP vaccine, there is a dramatic decline in the number of cases.

Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?
After the introduction of the vaccine, the number of cases declined dramatically. This is possibly due to more children getting vaccinated, hence the number of infected declines, hence the number of cases declines.

##Exploring CMI-PB data

Access data frame from the CMI-PB project

This database, like many modern projects, uses an API to return JSON format data.

We will use the R package 'jsonlite'.
```{r}
library(jsonlite)
 
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector= TRUE)
head(subject)
```
Q4. How many wP individuals and aP individuals are in this data set?
```{r}
table(subject$infancy_vac)
```
60 aP individuals and 58 wP individuals.

Q5. How many Male and Female subjects/patients are in the dataset?
```{r}
table(subject$biological_sex)
```
There are 79 female and 39 male subjects in this dataset

Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?
```{r}
table(subject$race, subject$biological_sex)
```
See above table for breakdown.

#Side-Note: Working with dates

Two of the columns of subject contain dates in the **Year-Month-Day** format. We can use the lubridate package to ease the pain of doing math with dates
```{r}
library(lubridate)

#Example for reference
today()
today() -ymd("2000-01-01")
today() -ymd("2002-5-15")
today() -mdy("05-15-2002")
```

Q8. Determine the age of all individuals at time of boost?

So what is the age of everyone in our dataset?
```{r}
subject$age <- time_length(today() -ymd(subject$year_of_birth), "years")

head(subject$age)
```
See above list.

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?
Yes, it looks like the wP lies farther right on the distribution and aP is farther left.


##Get more data from CMI-PB

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector=TRUE)
head(specimen)
```

##Joining Meta Tables

To know whether a given specimen_id comes from an aP or wP individual we need to link (a.k.a. “join” or merge) our specimen and subject data frames. The excellent dplyr package (that we have used previously) has a family of 'join()' functions that can help us with this common task.
```{r}
library(dplyr)

meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```

Now we can read some other data from the CMI-PB
```{r}
ab_titer <- read_json("https://www.cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE)

head(ab_titer)
```

One more 'inner_join()' to add all our metadata in 'meta' on to our 'ab_data' table:
```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```
```{r}
ggplot(abdata) +
  aes(MFI, antigen, col=infancy_vac) +
  geom_boxplot() +
  xlim(0, 2000) +
  theme_bw() +
  labs(x="Antigen", y="MFI", title = "Ab titer levels for all antigens")
```
Why are certain antigens and not others very variable in their detected levels here?
Likely due to vaccine differences.

```{r}
table(abdata$antigen)
```
There are potentially some difference here but in general it is hard to tell with this whole datset overview...

```{r}
table(abdata$dataset)
```

Lets focus in on just the 2021 dataset.
```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

